---
title: "Sea Level Time Series Prediction"
author: "Issac Chan, Vincent Paparang, Mikheil Parunovi, Denniz Zhang"
date: "2023-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading Packages
```{r Loading Packages, message=FALSE}
library(forecast)
library(zoo)
library(dplyr)
library(neuralnet)
```

Read in and clean Data
```{r}
global_sea_level <- read.csv("Team9.csv")
head(global_sea_level, 10)
clean_global_sea_level_1 <- global_sea_level[-1:-4, ]
colnames(clean_global_sea_level_1) <- c("Year", "TOPEX_Poseidon", "Jason_1", "Jason_2", "Jason_3")
clean_global_sea_level_1 <- clean_global_sea_level_1[-1, ]

clean_global_sea_level_1$Year <- as.numeric(clean_global_sea_level_1$Year)
clean_global_sea_level_1$Jason_3 <- as.numeric(clean_global_sea_level_1$Jason_3)
clean_global_sea_level_1$Jason_2 <- as.numeric(clean_global_sea_level_1$Jason_2)
clean_global_sea_level_1$Jason_1 <- as.numeric(clean_global_sea_level_1$Jason_1)
clean_global_sea_level_1$TOPEX_Poseidon <- as.numeric(clean_global_sea_level_1$TOPEX_Poseidon)

for (i in 1:nrow(clean_global_sea_level_1)){
  if (!is.na(clean_global_sea_level_1$Jason_3[i])){
    clean_global_sea_level_1$sea_level[i] <- clean_global_sea_level_1$Jason_3[i]
  } else if (!is.na(clean_global_sea_level_1$Jason_2[i])) {
    clean_global_sea_level_1$sea_level[i] <- clean_global_sea_level_1$Jason_2[i]
  } else if (!is.na(clean_global_sea_level_1$Jason_1[i])) {
    clean_global_sea_level_1$sea_level[i] <- clean_global_sea_level_1$Jason_1[i]
  } else if (!is.na(clean_global_sea_level_1$TOPEX_Poseidon[i])){
    clean_global_sea_level_1$sea_level[i] <- clean_global_sea_level_1$TOPEX_Poseidon[i]
  }
}

original_series <- clean_global_sea_level_1[, -which(names(clean_global_sea_level_1) %in% c("TOPEX_Poseidon", "Jason_1", "Jason_2", "Jason_3"))]

original.ts <- ts(original_series$sea_level, start = c(1993, 1), end = c(2022, 4), frequency = 36)

for (i in 1:nrow(clean_global_sea_level_1)){
  if(clean_global_sea_level_1$Year[i]%%1 <= 0.25){
    clean_global_sea_level_1$Quarter[i] <- paste0(floor(clean_global_sea_level_1$Year[i]), "Q1")
  } else if ((0.25 < clean_global_sea_level_1$Year[i]%%1) & (clean_global_sea_level_1$Year[i]%%1 <= 0.5)){
    clean_global_sea_level_1$Quarter[i] <- paste0(floor(clean_global_sea_level_1$Year[i]), "Q2")
  } else if ((0.5 < clean_global_sea_level_1$Year[i]%%1) & (clean_global_sea_level_1$Year[i]%%1 <= 0.75)){
    clean_global_sea_level_1$Quarter[i] <- paste0(floor(clean_global_sea_level_1$Year[i]), "Q3")
  } else if (0.75 < clean_global_sea_level_1$Year[i]%%1){
    clean_global_sea_level_1$Quarter[i] <- paste0(floor(clean_global_sea_level_1$Year[i]), "Q4")
  }
}

final_data <- data.frame(Quarter = clean_global_sea_level_1$Quarter, sea_level = clean_global_sea_level_1$sea_level)

final_quarterly_data <- final_data %>%
  group_by(Quarter) %>%
  summarise(avg = mean(sea_level)) %>%
  arrange(Quarter)

head(final_quarterly_data, 10)
```


Create time series model and visualize
```{r Original Data}
sea.ts <- ts(final_quarterly_data$avg, start = c(1993, 1), end = c(2022, 4), frequency = 4)

plot(original.ts, xlab = "Year", ylab = "Sea Level", main = "Sea Level Change - Original Series")

plot(sea.ts, xlab = "Year", ylab = "Sea Level", main = "Sea Level Change - Quarterly Series")
```


Detect if the data is a random walk with Acf, as well as check for the presence of seasonality in the series.
```{r Acf}
Acf(sea.ts)
Acf(diff(sea.ts))
```


Data Partition
```{r Data Partition}
nValid <- 16
nTrain <- length(sea.ts) - nValid
train.ts <- window(sea.ts, start = c(1993, 1), end = c(1993, nTrain))
valid.ts <- window(sea.ts, start = c(1993, nTrain + 1), end = c(1993, nTrain + nValid))
```


Linear Regression
```{r Linear}
linear.mod <- tslm(train.ts ~ trend)
summary(linear.mod)
```


Linear Regression with Season
```{r linear.season}
linear.season.mod <- tslm(train.ts ~ trend + season)
summary(linear.season.mod)
```


Linear Regression with Polynomial Trend and season
```{r Poly.season}
poly.season.mod <- tslm(train.ts ~ trend + I(trend^2) + season)
summary(poly.season.mod)
```


Holt-Winter Method
```{r Hwin}
hwin.mod <- ets(train.ts)
summary(hwin.mod)     # ETS (A,A,A) as suggested by the model
```


ARIMA Model
```{r ARIMA}
arima.mod <- auto.arima(train.ts)
summary(arima.mod)    # ARIMA(1,0,0)(2,1,0)[4] with drift as suggested by the model
```


Neural Network Model
```{r NN}
set.seed(13)
nn.mod <- nnetar(train.ts, P = 4, lambda = 1)
summary(nn.mod)
```


Tbats Model
```{r}
tbats.mod <- tbats(train.ts)
summary(tbats.mod)
```


Forecast Outcome
```{r Forecasting}
naive.mod.pred <- naive(train.ts, h = nValid, level = 0)
snaive.mod.pred <- snaive(train.ts, h = nValid, level = 0)
linear.mod.pred <- forecast(linear.mod, h = nValid, level = 0)
linear.season.mod.pred <- forecast(linear.season.mod, h = nValid, level = 0)
poly.season.mod.pred <- forecast(poly.season.mod, h = nValid, level = 0)
hwin.mod.pred <- forecast(hwin.mod, h = nValid, level = 0)
arima.mod.pred <- forecast(arima.mod, h = nValid, level = 0)
nn.mod.pred <- forecast(nn.mod, h = nValid, level = 0)
tbats.mod.pred <- forecast(tbats.mod, h = nValid, level = 0)
```


Accuracy for different models
```{r Accuracy, results='hide'}
accuracy(linear.mod.pred$mean, valid.ts)
accuracy(linear.season.mod.pred$mean, valid.ts)
accuracy(naive.mod.pred$mean, valid.ts)
accuracy(snaive.mod.pred$mean, valid.ts)
accuracy(poly.season.mod.pred$mean, valid.ts)
accuracy(hwin.mod.pred$mean, valid.ts)
accuracy(arima.mod.pred$mean, valid.ts)
accuracy(nn.mod.pred$mean, valid.ts)
accuracy(tbats.mod.pred$mean, valid.ts)
```


```{r Accuracy Table}
result_table <- rbind.data.frame(accuracy(linear.mod.pred$mean, valid.ts),
                                   accuracy(linear.season.mod.pred$mean, valid.ts),
                                   accuracy(naive.mod.pred$mean, valid.ts),
                                   accuracy(snaive.mod.pred$mean, valid.ts),
                                   accuracy(poly.season.mod.pred$mean, valid.ts),
                                   accuracy(hwin.mod.pred$mean, valid.ts),
                                   accuracy(arima.mod.pred$mean, valid.ts),
                                   accuracy(nn.mod.pred$mean, valid.ts),
                                   accuracy(tbats.mod.pred$mean, valid.ts))

rownames(result_table) <- c("linear.mod.pred",
                            "linear.season.mod.pred",
                            "naive.mod.pred",
                            "snaive.mod.pred",
                            "poly.season.mod.pred",
                            "hwin.mod.pred",
                            "arima.mod.pred",
                            "nn.mod.pred",
                            "tbats.mod.pred")

result_table
```


Plot for naive, seasonal naive and linear regression model
```{r}
plot(linear.mod.pred, include = 30, ylim = c(20,90), lwd = 2)
lines(naive.mod.pred$mean, col = "yellow", lwd = 2)
lines(snaive.mod.pred$mean, col = "orange", lwd = 2)
lines(valid.ts, col = "black", lwd = 2)
legend("topleft", legend = c("Linear Model", "Naive","Seasonal Naive", "Validation Data"), 
       col = c("skyblue", "yellow","orange", "black"), lty = 1, cex = 1)
```

Plot for linear regression with season and polynomial trend with season model
```{r}
plot(linear.season.mod.pred, include = 30, ylim = c(20,90), lwd = 2, col = "black", fcol = "red")
lines(poly.season.mod.pred$mean, col = "green", lwd = 2)
lines(valid.ts, col = "black", lwd = 2)
legend("topleft", legend = c("LR with season", "LR with Poly-Trend and Season", "Validation Data"), 
       col = c("red", "green", "black"), lty = 1, cex = 1)
```

Plot for Holt-Winter model
```{r}
plot(hwin.mod.pred, include = 30, ylim = c(20,90), lwd = 2, col = "black", fcol = "blue")
lines(valid.ts, col = "black", lwd = 2)
legend("topleft", legend = c("Holt-Winter", "Validation Data"), 
       col = c("blue", "black"), lty = 1, cex = 1)
```

Plot for ARIMA model
```{r}
plot(arima.mod.pred, include = 30, ylim = c(20,90), lwd = 2, col = "black", fcol = "pink")
lines(valid.ts, col = "black", lwd = 2)
legend("topleft", legend = c("ARIMA", "Validation Data"), 
       col = c("pink", "black"), lty = 1, cex = 1)
```

Plot for Neural Network model and 
```{r}
plot(nn.mod.pred, include = 30, ylim = c(20,90), lwd = 2, col = "black", fcol = "purple")
lines(tbats.mod.pred$mean, col = "brown", lwd = 2)
lines(valid.ts, col = "black", lwd = 2)
legend("topleft", legend = c("Neural Network", "TBATS", "Validation Data"), 
       col = c("purple", "brown", "black"), lty = 1, cex = 1)
```


Final plot for all predictions
```{r Plotting}
plot(linear.mod.pred, include = 30, ylim = c(20,90), lwd = 2)
lines(naive.mod.pred$mean, col = "yellow", lwd = 2)
lines(snaive.mod.pred$mean, col = "orange", lwd = 2)
lines(linear.season.mod.pred$mean, col = "red", lwd = 2)
lines(poly.season.mod.pred$mean, col = "green", lwd = 2)
lines(hwin.mod.pred$mean, col = "blue", lwd = 2)
lines(arima.mod.pred$mean, col = "pink", lwd = 2)
lines(nn.mod.pred$mean, col = "purple", lwd = 2)
lines(tbats.mod.pred$mean, col = "brown", lwd = 2)
lines(valid.ts, col = "black", lwd = 2)
legend("topleft", legend = c("Linear Model", "Seasonal Naive", "LR with season", "LR with Poly-Trend and Season", "Holt-Winter", "ARIMA", "Neural Network", "TBATS", "Validation Data"), 
       col = c("lightblue", "orange", "red", "green", "blue", "pink", "purple", "brown", "black"), lty = 1, cex = 0.7)
```


Future Prediction - Linear Regression with Polynomial Trend
```{r Future Forecasting 1}
full.poly.season.mod <- tslm(sea.ts ~ trend + I(trend^2) + season)
full.poly.season.mod.pred <- forecast(full.poly.season.mod, h = 20, level = 0)

plot(full.poly.season.mod.pred, include = 30, lwd = 2, col = "black", fcol = "green", main = "Forecast: Linear Regression with Polynomial Trend and Seasonality")
legend("topleft", legend = c("Prediction of Linear Model", "Historical Data"), col = c("green", "black"), lty = 1, cex = 0.8)

full.poly.season.mod.pred
```

Future Prediction - ARIMA
```{r Future Forecasting 2}
full.arima.mod <- Arima(sea.ts, order = c(1, 0, 0), seasonal = c(2, 1, 0))
full.arima.mod.pred <- forecast(full.arima.mod, h = 20, level = 0)

plot(full.arima.mod.pred, include = 30, lwd = 2, col = "black", fcol = "pink", main = "Forecast: ARIMA (1, 0, 0)(2, 1, 0)")
legend("topleft", legend = c("Prediction of ARIMA Model", "Historical Data"), col = c("pink", "black"), lty = 1, cex = 0.8)

full.arima.mod.pred
```

Future Prediction - Holt-Winter
```{r Future Forecasting 3}
full.hwin.mod <- ets(sea.ts, model = "AAA")
full.hwin.mod.pred <- forecast(full.hwin.mod, h = 20, level = 0)

plot(full.hwin.mod.pred, include = 30, lwd = 2, col = "black", fcol = "blue", main = "Forecast: Holt-Winter Smoothing Model")
legend("topleft", legend = c("Prediction of Holt-Winter Model", "Historical Data"), col = c("blue", "black"), lty = 1, cex = 0.8)

full.hwin.mod.pred
```

```{r}
Acf(full.poly.season.mod.pred$residuals)
Acf(full.arima.mod.pred$residuals)
Acf(full.hwin.mod.pred$residuals)
```

```{r Top 3 Future Prediction}
plot(full.poly.season.mod.pred, include = 30, lwd = 2, col = "black", fcol = "green")
lines(full.arima.mod.pred$mean, lwd = 2, col = "pink")
lines(full.hwin.mod.pred$mean, lwd = 2, col = "blue")
legend("topleft", legend = c("Future Prediction with linear regression\nwith polynomial trend with season", "Future Prediction with linear regression with ARIMA", "Future Prediction with linear regression Holt-Winter", "current"), col = c("green", "pink", "blue", "black"), lty = 1, cex = 0.8)
```

```
